---
title: "Week 11 — Adult Sex Ratio Estimation from Camera-Trap Data"
author: "SNR 690"
date: "Week 11"
format: html
categories: [camera traps, adult sex ratio, GAM, occupancy, Royle-Nichols, bootstrap]
---

# Week Overview

**Week 11 — Adult Sex Ratio Estimation: Comparing Naive, Predicted, and Occupancy-Based Approaches**

This week we explore three camera-trap–based approaches for estimating adult sex ratio (ASR) in red deer, using:

> Wohlfahrt, S., Edelhoff, H., Leitner, H., & Hackländer, K. (2026). Red deer adult sex ratio: comparing three approaches using camera trap data. *Journal of Wildlife Management* 90:e70139. <https://doi.org/10.1002/jwmg.70139>

| Session | Content | Duration |
|---------|---------|----------|
| **Session A** | Student paper presentation + ASR methods deep-dive + active learning | ~75 min |
| **Session B** | Hands-on R workshop: naive ASR, GAM prediction, RN occupancy | ~75 min |

---

::: {.callout-tip}
## Learning Objectives

By the end of this week, you will be able to:

1. **Define** adult sex ratio (ASR) and explain why accurate estimation matters for ungulate management and forest regeneration.
2. **Compare** three approaches to ASR estimation—naive, predicted (GAM), and occupancy-based (Royle–Nichols)—and articulate the assumptions, strengths, and failure modes of each.
3. **Implement** naive ASR calculation with bootstrap confidence intervals in R.
4. **Fit and evaluate** a GAM for sex prediction from camera-trap covariates, including model diagnostics and confusion matrix interpretation.
5. **Describe** Royle–Nichols occupancy modeling and explain how latent abundance estimates inform ASR.
6. **Critically evaluate** when simple methods (naive ASR) suffice versus when advanced approaches are necessary, using monitoring design constraints as criteria.
:::

---

# Before Class: Required Preparation

::: {.callout-important}
## Pre-Class Requirements

**Read:** Wohlfahrt et al. (2026) — full paper. Focus on the Methods and Results sections.

**Submit on Canvas before class** — written responses to:

1. **Application to your system:** Describe one specific scenario in your research (or a hypothetical one in your field) where estimating adult sex ratio from remote sensing data (camera traps, drone imagery, acoustic sensors) would be valuable. Name the species/taxon and explain why sex-specific detection bias could matter.

2. **Methods alignment:** Does the Wohlfahrt et al. (2026) paper demonstrate logical alignment from research objectives → data collection → statistical methods? Identify one strength and one weakness in 2–3 sentences.

3. **Question for the presenter:** Write one substantive question you want to ask about the paper's methods, results, or implications.
:::

**Guiding questions as you read:**

- What is the key equation for computing ASR? Under what assumption is the naive approach unbiased?
- Why did the authors choose weeks 9–17 as the comparison window? What ecological/statistical assumption does this satisfy?
- The GAM's prediction precision was ~75%. What does that mean in practical terms — how does imperfect prediction affect the predicted ASR?
- How does the Royle–Nichols model differ from a standard occupancy model? What additional parameter does it estimate?
- Under what conditions did the three approaches agree? When might they diverge?

---

# Session A — Paper Discussion & Methods Deep-Dive (~75 min)

## Schedule {.unnumbered}

| Time | Activity |
|------|----------|
| 0:00–0:05 | Entry ticket: retrieval practice |
| 0:05–0:15 | Instructor mini-lecture: ASR concepts and three approaches |
| 0:15–0:31 | Student paper presentation + guided discussion |
| 0:31–0:46 | Jigsaw activity: become an expert in one ASR approach |
| 0:46–0:56 | Choose-your-own-method scenario + gallery walk |
| 0:56–1:06 | Structured mini-debate |
| 1:06–1:12 | Concept map + muddiest point |
| 1:12–1:15 | Minute paper |

---

## Entry Ticket — 0:00–0:05 {.unnumbered}

On your notecard, write brief answers to:

1. What is the formula for adult sex ratio (ASR)? Write it without looking at your notes.
2. From Week 10: What is the core problem that occupancy models solve? (one sentence)
3. If 40% of your camera-trap sightings cannot be sexed, how might that bias your ASR estimate? In which direction?

---

## Key Concepts: ASR and Three Approaches — 0:05–0:15 {.unnumbered}

**Adult sex ratio (ASR)** is a fundamental demographic parameter:

$$\text{ASR} = \frac{N_{\text{males}}}{N_{\text{males}} + N_{\text{females}}}$$

The paper compares three approaches to estimate ASR from camera-trap data:

| Approach | What it does | Key assumption |
|----------|-------------|----------------|
| **Naive** | Use only sex-identified sightings; exclude unknowns | Unknowns are missing completely at random (MCAR) |
| **Predicted** | Use a GAM to predict sex for unknown sightings, then recompute ASR | GAM correctly classifies sex; prediction errors are balanced |
| **Occupancy (RN)** | Royle–Nichols model estimates latent abundance by sex, accounting for detection probability | Population closure during sampling window; constant per-individual detection |

**Study snapshot:** 24 cameras · 20,250 camera-days · 4,110 adult/subadult sightings · 436 unknowns (~8.9%) · Austrian Alps · Jan 2020–May 2022 · Focal weeks 9–17

---

## Student Paper Presentation — 0:15–0:31 {.unnumbered}

**Your classmate will present Wohlfahrt et al. (2026).** As you listen:

- Note which ASR approach you find most convincing and why
- Prepare one follow-up question or observation

### Rotating Discussion Roles {.unnumbered}

Today's assigned roles:

| Role | Your job |
|------|----------|
| **Facilitator** | Keep discussion flowing; call on quiet voices |
| **Skeptic** | Challenge assumptions; play devil's advocate |
| **Scribe** | Capture key points on the whiteboard |
| **Assumption-Checker** | Flag statistical assumptions being made or violated |

*Everyone else: contribute to discussion freely.*

### Presentation Template (for the presenter) {.unnumbered}

::: {.callout-note}
## What to Cover in Your Presentation (~8 min)

1. **Motivation** (~1 min): Why does ASR matter for red deer management? Why are camera traps useful but tricky?
2. **Data** (~1.5 min): Describe the study area, camera deployment (24 cameras, trail-based), sighting classification (species, sex, age), and the unknown-sex problem.
3. **Three Approaches** (~3.5 min):
   - Naive ASR: formula, what's excluded, when it works
   - Predicted ASR: GAM model, covariates used, prediction precision (~75%), threshold (0.5)
   - Occupancy RN: detection histories, latent abundance, how ASR is derived from $\bar{N}$ estimates
4. **Key Results** (~1.5 min): ASR estimates across years and methods (with CIs), Wilcoxon tests, the main finding (methods agreed under these conditions)
5. **Critique** (~0.5 min): One limitation you identified; one thing you would do differently

**Prepare 2 discussion questions** to pose to the class after your talk.
:::

---

## Jigsaw Activity: Three ASR Approaches — 0:31–0:46 {.unnumbered}

::: {.callout-note}
## Become an Expert, Then Teach

**Expert phase (7 min):** You'll be assigned to one method. With your partner(s):

1. Write the key equation(s) for your method on the poster/whiteboard
2. List 2–3 assumptions your method requires
3. Identify 1–2 scenarios where your method would fail or give biased results
4. Prepare a 2-minute teach-back

| Group | Method | Key focus |
|-------|--------|-----------|
| **A** | Naive ASR | $\text{ASR} = N_m / (N_m + N_f)$; when does excluding unknowns bias estimates? |
| **B** | Predicted ASR (GAM) | How does the GAM predict sex? What happens at the 0.5 threshold? |
| **C** | Occupancy RN | $P(\text{detect} \geq 1 \mid N, r) = 1 - (1-r)^N$; what is latent abundance? |

**Teach-back phase (7 min):** Regroup into mixed teams (one person from each group). Each expert teaches their method in ~2 min. Ask questions!
:::

---

## Choose-Your-Own-Method Scenario — 0:46–0:56 {.unnumbered}

Your group will receive a wildlife monitoring scenario. Decide:

- Which ASR approach(es) you would recommend
- Why that approach is the best fit for the constraints described
- What risks or limitations remain

Write your recommendation on your poster. Then do a **gallery walk** — visit other groups' posters and add comments.

---

## Structured Mini-Debate — 0:56–1:06 {.unnumbered}

**Motion:** *"Naive ASR is sufficient for most wildlife management applications."*

- **Side A — "Keep It Simple":** Argue for naive ASR
- **Side B — "Model It Right":** Argue for occupancy-based or predicted ASR

Each side prepares arguments (2 min), then presents (2 min each), followed by open rebuttal (2 min) and synthesis (2 min).

---

## Concept Map + Muddiest Point — 1:06–1:12 {.unnumbered}

On your notecard:

1. Draw a quick concept map linking: **ASR, naive, predicted, GAM, occupancy, RN, detection probability, latent abundance, bootstrap CI, camera placement, unknowns**. Connect terms with labeled arrows (e.g., "assumes," "accounts for," "biased by").
2. Write your **muddiest point**: the one concept from today that is least clear.
3. Discuss with a neighbor — can they clarify?

---

## Minute Paper — 1:12–1:15 {.unnumbered}

On your notecard:

1. What is the single most important insight about ASR estimation from today?
2. Which of the three approaches would you use in your own research system, and why?
3. One question you still have heading into the workshop.

---

# Session B — Workshop: ASR Estimation in R (~75 min)

## Schedule {.unnumbered}

| Time | Activity |
|------|----------|
| 0:00–0:05 | Setup + debrief muddiest points from Session A |
| 0:05–0:25 | Exercise A: Naive ASR + bootstrap CI |
| 0:25–0:50 | Exercise B: GAM sex prediction + evaluation |
| 0:50–1:05 | Exercise C: RN occupancy ASR |
| 1:05–1:10 | Peer comparison |
| 1:10–1:15 | Wrap-up + next steps |

## Setup {.unnumbered}

```{r}
#| label: setup-packages
#| echo: true
#| eval: false

# Install if needed
# install.packages(c("mgcv", "boot", "unmarked", "dplyr", "ggplot2", "tidyr"))

library(mgcv)
library(boot)
library(unmarked)
library(dplyr)
library(ggplot2)

set.seed(690)
```

---

## Exercise A: Naive ASR + Bootstrap CI (20 min) {.unnumbered}

**Goal:** Compute weekly ASR from sex-known sightings (weeks 9–17) and bootstrap a confidence interval.

### Step 1: Simulate data {.unnumbered}

```{r}
#| label: simulate-data
#| echo: true
#| eval: false

set.seed(690)

# Simulated camera-trap sightings inspired by Wohlfahrt et al. (2026)
n_sightings <- 4110
n_cameras <- 24
weeks <- 9:17

sightings <- data.frame(
  sighting_id = 1:n_sightings,
  camera_id = factor(sample(1:n_cameras, n_sightings, replace = TRUE)),
  week = sample(weeks, n_sightings, replace = TRUE),
  day_of_year = sample(60:120, n_sightings, replace = TRUE),
  hour = sample(0:23, n_sightings, replace = TRUE),
  age_class = sample(c("adult", "subadult", "juvenile"),
                     n_sightings, replace = TRUE,
                     prob = c(0.65, 0.20, 0.15)),
  distance_to_road = runif(n_sightings, 50, 2000)
)

# Assign sex: ~8.9% unknown; males slightly less common (ASR ~ 0.42)
sightings$sex <- sample(c("male", "female", "unknown"),
                        n_sightings, replace = TRUE,
                        prob = c(0.38, 0.53, 0.089))
```

### Step 2: Filter and compute weekly naive ASR {.unnumbered}

```{r}
#| label: naive-asr
#| echo: true
#| eval: false

# Filter: adults and subadults only (exclude juveniles!)
adults <- sightings %>%
  filter(age_class %in% c("adult", "subadult"))

# Filter: sex-known only (naive approach excludes unknowns)
sex_known <- adults %>%
  filter(sex %in% c("male", "female"))

# Compute weekly ASR
weekly_asr <- sex_known %>%
  group_by(week) %>%
  summarise(
    n_males = sum(sex == "male"),
    n_females = sum(sex == "female"),
    asr = n_males / (n_males + n_females),
    .groups = "drop"
  )

print(weekly_asr)
cat("Mean naive ASR:", mean(weekly_asr$asr), "\n")
```

### Step 3: Bootstrap CI {.unnumbered}

```{r}
#| label: bootstrap-naive
#| echo: true
#| eval: false

# Bootstrap mean ASR across weeks (R = 1000)
boot_mean_asr <- function(data, indices) {
  mean(data$asr[indices])
}

boot_result <- boot(data = weekly_asr,
                    statistic = boot_mean_asr,
                    R = 1000)

boot_ci <- boot.ci(boot_result, type = "perc")
cat("Bootstrap 95% CI:", boot_ci$percent[4], "-", boot_ci$percent[5], "\n")
```

::: {.callout-tip}
## Checkpoint A

**Expected output:** Mean ASR around 0.38–0.44; CI width of approximately ±0.03–0.06.

**Common pitfalls:**

- Forgot to exclude juveniles → ASR will be wrong
- Bootstrapping individual sightings instead of weekly ASRs → wrong unit of resampling
- Getting `NaN` → a week has 0 sex-known sightings after filtering
:::

---

## Exercise B: GAM Sex Prediction + Evaluation (25 min) {.unnumbered}

**Goal:** Fit a GAM to predict sex from covariates, evaluate it, then use predictions for unknown sightings to compute predicted ASR.

### Step 1: Train/test split {.unnumbered}

```{r}
#| label: gam-split
#| echo: true
#| eval: false

set.seed(690)

sex_known$sex_binary <- ifelse(sex_known$sex == "male", 1, 0)

# 80/20 split
train_idx <- sample(1:nrow(sex_known), size = 0.8 * nrow(sex_known))
train <- sex_known[train_idx, ]
test  <- sex_known[-train_idx, ]
```

### Step 2: Fit GAM {.unnumbered}

```{r}
#| label: gam-fit
#| echo: true
#| eval: false

# GAM with smooths for day_of_year and hour,
# camera_id as random effect (accounts for camera-specific detection heterogeneity),
# interaction: distance_to_road × day_of_year
gam_model <- gam(
  sex_binary ~ s(day_of_year, bs = "cr") +
               s(hour, bs = "cr") +
               s(camera_id, bs = "re") +
               ti(distance_to_road, day_of_year, bs = "cr"),
  data = train,
  family = binomial(link = "logit"),
  method = "REML"
)

summary(gam_model)
```

### Step 3: Evaluate {.unnumbered}

```{r}
#| label: gam-eval
#| echo: true
#| eval: false

# Predict on test set
test$pred_prob <- predict(gam_model, newdata = test, type = "response")
test$pred_sex <- ifelse(test$pred_prob > 0.5, "male", "female")

# Confusion matrix
conf_mat <- table(Predicted = test$pred_sex, Actual = test$sex)
print(conf_mat)

accuracy <- sum(diag(conf_mat)) / sum(conf_mat)
cat("Accuracy:", round(accuracy, 3), "\n")
```

### Step 4: Predict unknowns → recompute ASR {.unnumbered}

```{r}
#| label: predicted-asr
#| echo: true
#| eval: false

# Predict sex for unknown sightings
unknowns <- adults %>% filter(sex == "unknown")
unknowns$pred_prob <- predict(gam_model, newdata = unknowns, type = "response")
unknowns$sex <- ifelse(unknowns$pred_prob > 0.5, "male", "female")

# Combine and recompute weekly ASR
combined <- bind_rows(
  sex_known %>% select(week, sex),
  unknowns %>% select(week, sex)
)

weekly_asr_predicted <- combined %>%
  group_by(week) %>%
  summarise(
    n_males = sum(sex == "male"),
    n_females = sum(sex == "female"),
    asr = n_males / (n_males + n_females),
    .groups = "drop"
  )

# Bootstrap
boot_result_pred <- boot(data = weekly_asr_predicted,
                         statistic = boot_mean_asr,
                         R = 1000)
boot_ci_pred <- boot.ci(boot_result_pred, type = "perc")

cat("Predicted ASR mean:", mean(weekly_asr_predicted$asr), "\n")
cat("95% CI:", boot_ci_pred$percent[4], "-", boot_ci_pred$percent[5], "\n")
```

::: {.callout-tip}
## Checkpoint B

**Expected output:** Predicted ASR should be similar to naive ASR (since unknowns are only ~8.9% of data in this simulation). The CI may be slightly narrower because more data is included.

**Common pitfalls:**

- Forgot `set.seed()` before train/test split
- `camera_id` not converted to factor → GAM treats it as continuous
- Confusion matrix dimensions don't match → check factor levels
:::

---

## Exercise C: Royle–Nichols Occupancy ASR (15 min) {.unnumbered}

**Goal:** Build detection histories and fit a Royle–Nichols model for one sex. Understand how latent abundance translates to ASR.

### Conceptual Framework {.unnumbered}

The Royle–Nichols (RN) model estimates **latent abundance** ($\lambda$) from detection/non-detection data:

$$P(\text{detect} \geq 1 \mid N, r) = 1 - (1-r)^N$$

where $r$ is per-individual detection probability and $N \sim \text{Poisson}(\lambda)$.

For ASR: fit RN separately for males and females, then:

$$\text{ASR}_{\text{occupancy}} = \frac{\bar{N}_{\text{male}}}{\bar{N}_{\text{male}} + \bar{N}_{\text{female}}}$$

### Step 1: Build detection histories {.unnumbered}

```{r}
#| label: detection-histories
#| echo: true
#| eval: false

library(tidyr)

# Male detection histories: cameras × weeks
males_only <- adults %>% filter(sex == "male")

det_hist_male <- males_only %>%
  mutate(detected = 1) %>%
  pivot_wider(
    id_cols = camera_id,
    names_from = week,
    values_from = detected,
    values_fn = ~1,
    values_fill = 0
  ) %>%
  arrange(camera_id)

# Ensure all weeks 9-17 are present
for (w in 9:17) {
  if (!(as.character(w) %in% names(det_hist_male))) {
    det_hist_male[[as.character(w)]] <- 0
  }
}

det_matrix_male <- as.matrix(det_hist_male[, as.character(9:17)])
head(det_matrix_male)
```

### Step 2: Fit RN model {.unnumbered}

```{r}
#| label: rn-fit
#| echo: true
#| eval: false

library(unmarked)

umf_male <- unmarkedFrameOccu(y = det_matrix_male)

# Royle-Nichols model (intercept-only for simplicity)
rn_male <- occuRN(~1, ~1, data = umf_male)
summary(rn_male)

lambda_male <- exp(coef(rn_male, type = "state"))
cat("Estimated mean latent abundance (males):", lambda_male, "\n")
```

### Step 3: Compute occupancy-based ASR {.unnumbered}

```{r}
#| label: rn-asr
#| echo: true
#| eval: false

# YOUR TASK: Repeat Steps 1-2 for females
# Then compute:
# asr_occupancy <- lambda_male / (lambda_male + lambda_female)

# Compare all three ASR estimates:
# cat("Naive ASR:", mean(weekly_asr$asr), "\n")
# cat("Predicted ASR:", mean(weekly_asr_predicted$asr), "\n")
# cat("Occupancy ASR:", asr_occupancy, "\n")
```

::: {.callout-tip}
## Checkpoint C

**Expected output:** Lambda estimates should be positive. Occupancy-based ASR should be in a plausible range (0.3–0.6).

**Key things to check:**

- Detection matrix dimensions: should be (number of cameras) × 9
- No extra columns (remove `camera_id` before converting to matrix)
- Model converges without warnings
:::

---

## Peer Comparison — 1:05–1:10 {.unnumbered}

Pair up and compare your results:

1. Your naive ASR estimate and CI
2. Your predicted ASR estimate and CI
3. Your RN lambda estimate

**Discussion:** Do your three estimates agree? What explains any differences?

---

## Wrap-Up — 1:10–1:15 {.unnumbered}

::: {.callout-important}
## What to Submit / What to Bring Next Time

**Submit (via Canvas, if applicable):**

- Your completed R script or `.qmd` with Exercises A–C
- A 2–3 sentence reflection: Which ASR approach would you use for your own research system, and why?

**Bring next time:**

- Progress on your course project (be ready for a brief update if called upon)
- One remaining question about occupancy or ASR methods
:::

---

## Key Results from the Paper (Reference) {.unnumbered}

For comparing your simulation results to the actual paper findings:

| Year | Naive ASR (95% CI) | Predicted ASR (95% CI) | Occupancy ASR (95% CI) |
|------|--------------------|------------------------|------------------------|
| 2020 | 0.40 (0.34–0.46) | 0.44 (0.39–0.49) | 0.40 (0.34–0.47) |
| 2021 | 0.40 (0.31–0.49) | 0.46 (0.39–0.54) | 0.36 (0.36–0.36) |
| 2022 | 0.42 (0.35–0.50) | 0.48 (0.41–0.56) | 0.45 (0.44–0.46) |

*Source: Wohlfahrt et al. (2026), Table/Figure for weeks 9–17.*

**Key finding:** Under these study conditions (~8.9% unknowns, trail-based cameras), all three methods produced comparable ASR estimates with overlapping CIs. No statistically significant differences (Wilcoxon signed-rank tests, $p > 0.05$).

---
