---
title: "Week 9 - Split-Plot Demo"
subtitle: "SNR 690 - Contemporary Approaches to Quantitative Methods"
author: "For instructor demo"
format:
  revealjs:
    theme: default
    slide-number: true
    incremental: false
    scrollable: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Split-Plot Designs: Quick Demo

**Purpose:** 5-8 minute instructor-led walkthrough of split-plot design concepts and R code.

**Reading:** Adcock et al. (1999) - No-tillage culture and nitrogen fertilizer management for burley tobacco production.

<https://www.cambridge.org/core/journals/journal-of-agricultural-science/article/notillage-culture-and-nitrogen-fertilizer-management-for-burley-tobacco-production/C7BCABF749500FB9F69D06DF444D4F10>

---

## The Cambridge Paper at a Glance

- **Goal:** Compare tillage methods (conventional vs. no-till) and nitrogen fertilizer rates on burley tobacco yield
- **Design:** Split-plot with tillage as the main-plot factor and fertilizer as the subplot factor
- **Main plots:** Tillage (CT, NT) - applied to whole plots because tillage requires large equipment
- **Subplots:** Fertilizer (0, 140, 280 kg N/ha) - can vary within a main plot
- **Replicates:** 4 blocks (randomized complete block design at the main-plot level)

---

## Split-Plot Design Diagram

```{r}
#| label: fig-splitplot-diagram
#| fig-cap: "Schematic of a split-plot design with 2 tillage levels and 3 fertilizer rates across 4 blocks."
#| fig-width: 9
#| fig-height: 5
#| echo: true

library(ggplot2)

# Build a schematic of the split-plot layout
blocks <- rep(1:4, each = 6)
tillage <- rep(rep(c("CT", "NT"), each = 3), 4)
fertilizer <- rep(c("0", "140", "280"), 8)

diagram_data <- data.frame(
  Block = factor(blocks),
  Tillage = tillage,
  Fertilizer = fertilizer
)

# Create x and y positions for the grid
diagram_data$x <- as.numeric(diagram_data$Block)
diagram_data$y <- ifelse(diagram_data$Tillage == "CT", 2, 1)
fert_offset <- ifelse(diagram_data$Fertilizer == "0", -0.25,
               ifelse(diagram_data$Fertilizer == "140", 0, 0.25))
diagram_data$x <- diagram_data$x + fert_offset

diagram_data$label <- paste0(diagram_data$Tillage, "\n", diagram_data$Fertilizer, " kg N")

ggplot(diagram_data, aes(x = x, y = y, fill = Fertilizer)) +
  geom_tile(width = 0.22, height = 0.6, color = "black") +
  geom_text(aes(label = label), size = 2.3) +
  scale_fill_manual(
    values = c("0" = "#a6d96a", "140" = "#fdae61", "280" = "#d73027"),
    name = "Fertilizer\n(kg N/ha)"
  ) +
  scale_y_continuous(
    breaks = c(1, 2),
    labels = c("NT (No-Till)", "CT (Conventional)"),
    limits = c(0.3, 2.7)
  ) +
  scale_x_continuous(
    breaks = 1:4,
    labels = paste("Block", 1:4)
  ) +
  labs(
    title = "Split-Plot Layout",
    x = "Block (Replicate)",
    y = "Main Plot (Tillage)"
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank())
```

---

## Key Properties of Split-Plot Designs

- **Two levels of randomization:**
  - Main-plot factor (tillage) is randomized across whole plots within each block
  - Subplot factor (fertilizer) is randomized within each main plot
- **Two error terms:**
  - Main-plot error: variation among main plots within blocks (used to test tillage)
  - Subplot error: variation among subplots within main plots (used to test fertilizer and the interaction)
- **Why split-plot?** Some factors (e.g., tillage) are hard or impractical to apply at a small scale

---

## The Model

In R, we fit the split-plot model using `lmer` from the `lme4` package:

```r
# Yield ~ fixed effects + random effects
# Tillage * Fertilizer = Tillage + Fertilizer + Tillage:Fertilizer
# (1 | Block/WU) = random intercept for Block and whole-plot unit nested in Block

model <- lmer(Yield ~ Tillage * Fertilizer + (1 | Block/WU), data = data)
```

| Term | Meaning |
|------|---------|
| `Tillage * Fertilizer` | Fixed effects: main effects and interaction |
| `(1 | Block)` | Random effect for blocks (replicates) |
| `(1 | Block:WU)` | Random effect for whole-plot units (main-plot error) |
| Residual | Subplot error |

---

## Fitting the Model in R

```{r}
#| label: fit-splitplot-model
#| echo: true
#| eval: true

# Source the simulation and plotting code from the repo.
# This script (projects/examplesplitplot.R) creates:
#   - data: a data.frame with Tillage, Fertilizer, Replicate, Yield, WU, Block
#   - model: an lmer fit (Yield ~ Tillage * Fertilizer + (1 | Block/WU))
#   - p: a ggplot object showing the simulated split-plot data
source("projects/examplesplitplot.R")

# Display the ANOVA table
anova(model)
```

---

## Model Summary

```{r}
#| label: model-summary
#| echo: true
#| eval: true

summary(model)
```

---

## Diagnostic Plot: Residuals vs Fitted

```{r}
#| label: fig-diagnostics
#| fig-cap: "Residuals vs fitted values for the split-plot model. Look for patterns indicating model misspecification."
#| fig-width: 7
#| fig-height: 4.5
#| echo: true
#| eval: true

# Diagnostic plot: residuals vs fitted values
diag_data <- data.frame(
  Fitted = fitted(model),
  Residuals = residuals(model)
)

ggplot(diag_data, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 14)
```

---

## Interpreting the Results

- **Main-plot effect (Tillage):** Tested against the whole-plot error. A significant tillage effect means the two tillage methods produce different average yields.
- **Subplot effect (Fertilizer):** Tested against the subplot (residual) error. A significant fertilizer effect means nitrogen rate influences yield.
- **Interaction (Tillage x Fertilizer):** Also tested against subplot error. A significant interaction means the fertilizer effect depends on which tillage method is used.
- **Check:** The whole-plot error variance should be non-negligible. If it is near zero, the split-plot structure may not be needed.

---

## Practical Tips

- **Balance:** Ensure equal replication of all treatment combinations within each block.
- **Blocking:** Use blocks to control for known sources of variation (e.g., field position, soil type).
- **Diagnostics:** Always check residual plots for non-constant variance or non-normality.
- **Random vs. fixed:** The whole-plot grouping factor must be random. If you treat it as fixed, you underestimate the main-plot error.
- **Software:** `lme4::lmer` and `nlme::lme` both handle split-plot models. Use `car::Anova()` or `anova()` for Type III tests.

---

## Next Steps

**Student activity (10 min):**

1. Write a split-plot experiment proposal for your thesis or research area.
2. Identify your main-plot factor and subplot factor.
3. Describe how you would analyze the data (e.g., write out the `lmer` call).

Share with a neighbor and be ready to discuss.

---

## References

- **Cambridge paper:** Adcock, C. W., Farthing, G. R., & Williams, J. D. (1999). No-tillage culture and nitrogen fertilizer management for burley tobacco production. *Journal of Agricultural Science*. <https://www.cambridge.org/core/journals/journal-of-agricultural-science/article/notillage-culture-and-nitrogen-fertilizer-management-for-burley-tobacco-production/C7BCABF749500FB9F69D06DF444D4F10>
- **R code:** `projects/examplesplitplot.R` - source with `source("projects/examplesplitplot.R")`
- **Packages:** `lme4`, `car`, `ggplot2`, `dplyr`

```{r}
#| label: install-note
#| eval: false
#| echo: true
# Install required packages if needed:
install.packages(c("lme4", "car", "ggplot2", "dplyr"))
```
