---
title: "Week 03 Session B — Answer Sheet"
format: 
  html:
    toc: true
    toc-depth: 3
    include-in-header:
      text: |
        <style>
          #protected-content { display: none; }
          #password-form { 
            max-width: 450px; 
            margin: 100px auto; 
            padding: 40px;
            border: 2px solid #dc3545;
            border-radius: 12px;
            text-align: center;
            background-color: #fff8f8;
          }
          #password-form h2 {
            color: #dc3545;
            margin-bottom: 20px;
          }
          #password-form input[type="password"] {
            padding: 12px;
            width: 250px;
            margin: 15px 0;
            font-size: 16px;
            border: 1px solid #ccc;
            border-radius: 6px;
          }
          #password-form button {
            padding: 12px 30px;
            font-size: 16px;
            cursor: pointer;
            background-color: #dc3545;
            color: white;
            border: none;
            border-radius: 6px;
            transition: background-color 0.3s;
          }
          #password-form button:hover {
            background-color: #c82333;
          }
          .error-message {
            color: #dc3545;
            margin-top: 15px;
            font-weight: bold;
          }
          .warning-text {
            color: #666;
            font-size: 14px;
            margin-top: 20px;
          }
        </style>
---

<div id="password-form">
  <h2>🔒 Instructor Access Only</h2>
  <p>This page contains solutions to the Week 03 coding activities.</p>
  <p><strong>Students:</strong> Please work through the problems yourself before viewing answers!</p>
  <input type="password" id="password-input" placeholder="Enter instructor password">
  <br>
  <button onclick="checkPassword()">Access Answers</button>
  <p id="error-msg" class="error-message"></p>
  <p class="warning-text">If you are a student and don't have the password,<br>that means you shouldn't be here yet! 😊</p>
</div>

<div id="protected-content">

::: {.callout-warning}
## Instructor Use Only
This document contains solutions and discussion points for the differentiated coding activities. Do not share with students until after the session.
:::

---

# 🟢 Track A: Beginners

## Option A2: My First Model — Expected Answers

### Question: What does the slope for temperature mean?

**Expected answer:** For every 1-degree increase in temperature, yield increases by approximately 1.3 units (the exact value depends on the data). This assumes a linear relationship and that all other factors are held constant.

### Question: Which model explains more variance?

**Expected answer:** Model 2 (with two predictors) will typically have a higher R-squared than Model 1. However, students should note:

- R-squared always increases when you add predictors
- Adjusted R-squared accounts for the number of predictors
- A higher R-squared doesn't always mean a "better" model

### Question: Interpret the coefficient for rainfall while "controlling for" temperature

**Expected answer:** Holding temperature constant, for every 1-unit increase in rainfall, yield changes by [X] units. This is the *partial* effect of rainfall — the effect of rainfall after accounting for the relationship between temperature and yield.

**Discussion point:** With the provided data, temperature and rainfall are correlated. This can lead to:

- Coefficients that differ from simple correlations
- Larger standard errors (multicollinearity)
- Difficulty interpreting "independent" effects

---

# 🟡 Track B: Intermediate Stations

## Station 1: Salamander Survival — Answer

### What's wrong with the proposed method?

The proposed method `glm(survival ~ canopy_cover, family = binomial)` treats each salamander as an independent observation. But salamanders within the same forest fragment are **not independent** - they share the same environment, microclimate, predators, etc.

**The problem:** Pseudoreplication. The true sample size is 12 fragments, not 240 salamanders.

### The correct approach

Use a **generalized linear mixed model (GLMM)** with fragment as a random effect:

```{r}
#| label: station1-answer
#| eval: false

library(tidyverse)
library(lme4)

# -------------------------------------------
# Simulate the data
# -------------------------------------------

set.seed(123)

n_fragments <- 12
n_salamanders_per_fragment <- 20

# Create fragment-level data
fragments <- data.frame(
  fragment_id = 1:n_fragments,
  canopy_cover = runif(n_fragments, 30, 90)
)

# Add fragment-level random effect
# This represents unmeasured fragment-level variation
fragments$fragment_effect <- rnorm(n_fragments, 0, 0.8)

# Create individual salamander data
salamander_data <- fragments %>%
  slice(rep(1:n(), each = n_salamanders_per_fragment)) %>%
  mutate(
    salamander_id = 1:n(),
    # True model: logit(survival) = -2 + 0.04*canopy + fragment_effect
    log_odds = -2 + 0.04 * canopy_cover + fragment_effect,
    prob_survival = plogis(log_odds),
    survival = rbinom(n(), 1, prob_survival)
  )

# -------------------------------------------
# WRONG approach (ignoring clustering)
# -------------------------------------------

wrong_model <- glm(survival ~ canopy_cover, 
                   family = binomial, 
                   data = salamander_data)
summary(wrong_model)

# -------------------------------------------
# CORRECT approach (mixed model)
# -------------------------------------------

correct_model <- glmer(survival ~ canopy_cover + (1 | fragment_id), 
                       family = binomial, 
                       data = salamander_data)
summary(correct_model)

# -------------------------------------------
# Compare results
# -------------------------------------------

# Extract coefficients and SEs
wrong_coef <- summary(wrong_model)$coefficients
correct_coef <- summary(correct_model)$coefficients

cat("WRONG model SE for canopy_cover:", wrong_coef["canopy_cover", "Std. Error"], "\n")
cat("CORRECT model SE for canopy_cover:", correct_coef["canopy_cover", "Std. Error"], "\n")
```

### What students should observe

| Comparison | Wrong Model | Correct Model |
|------------|-------------|---------------|
| Standard error for canopy_cover | Smaller (artificially precise) | Larger (realistic) |
| p-value | Often significant | May not be significant |
| Conclusion | May claim strong effect | More honest uncertainty |

**Key insight:** Ignoring the clustering leads to **overconfident conclusions** because we're pretending we have 240 independent observations when we really only have 12 independent units.

---

## Station 4: Soil Microbial Diversity — Answer

### Is Kruskal-Wallis wrong, or just overly cautious?

Kruskal-Wallis is **not wrong**, but it may be **unnecessarily conservative** for this scenario.

**When Kruskal-Wallis is truly needed:**

- Severely non-normal data (heavy skew, outliers)
- Ordinal data
- Very small sample sizes where normality can't be assessed
- Heterogeneous variances that can't be addressed

**When ANOVA is fine:**

- Slight skewness (ANOVA is robust to this)
- Moderate sample sizes (n ≥ 10 per group)
- Residuals approximately normal
- Similar variances across groups

### The comparison

```{r}
#| label: station4-answer
#| eval: false

library(tidyverse)

# -------------------------------------------
# Simulate the data
# -------------------------------------------

set.seed(456)

n_per_treatment <- 10

soil_data <- data.frame(
  treatment = rep(c("no_till", "reduced_till", "conventional"), each = n_per_treatment),
  field_id = 1:30
)

# Simulate slightly right-skewed Shannon diversity
# True effect: no-till > reduced > conventional
soil_data <- soil_data %>%
  mutate(
    true_mean = case_when(
      treatment == "no_till" ~ 3.8,
      treatment == "reduced_till" ~ 3.2,
      treatment == "conventional" ~ 2.6
    ),
    # Add slight right skew using gamma-like distribution
    diversity = true_mean + rnorm(n(), 0, 0.5) + rexp(n(), rate = 5)
  )

# Visualize
ggplot(soil_data, aes(x = treatment, y = diversity)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5)

# -------------------------------------------
# Proposed: Kruskal-Wallis
# -------------------------------------------

kruskal_result <- kruskal.test(diversity ~ treatment, data = soil_data)
print(kruskal_result)

# -------------------------------------------
# Alternative: ANOVA
# -------------------------------------------

anova_model <- aov(diversity ~ treatment, data = soil_data)
summary(anova_model)

# Check assumptions
par(mfrow = c(2, 2))
plot(anova_model)

# Post-hoc (only available with ANOVA, not Kruskal-Wallis easily)
TukeyHSD(anova_model)
```

### What students should observe

| Aspect | Kruskal-Wallis | ANOVA |
|--------|----------------|-------|
| Tests | Differences in ranks | Differences in means |
| Interpretation | Less intuitive | Effect sizes in original units |
| Post-hoc tests | Dunn's test (less common) | Tukey HSD (standard) |
| Power | Slightly lower | Slightly higher |

**Key insight:** For slightly skewed but otherwise well-behaved data, ANOVA is often the better choice because:

1. It's more powerful
2. Effect sizes are interpretable
3. Post-hoc comparisons are straightforward
4. ANOVA is robust to moderate violations of normality

---

## Station 6: Seedling Germination — Answer

### What's wrong with `lm(proportion ~ stratification_time)`?

Three major problems:

1. **Bounded response:** Proportions are bounded between 0 and 1. Linear regression can predict values outside this range.

2. **Non-constant variance:** Variance of proportions depends on the mean. Near 0 or 1, variance is smaller. This violates homoscedasticity.

3. **Ignores sample size:** A proportion of 0.5 from 10 seeds has more uncertainty than 0.5 from 100 seeds. Linear regression treats them the same.

### The correct approach

Use **binomial GLM with counts** (successes and failures):

```{r}
#| label: station6-answer
#| eval: false

library(tidyverse)

# -------------------------------------------
# Simulate the data
# -------------------------------------------

set.seed(789)

n_dishes <- 10
n_seeds <- 50
strat_times <- c(0, 30, 60, 90)

germination_data <- expand.grid(
  dish_id = 1:n_dishes,
  stratification_time = strat_times
) %>%
  mutate(
    # True relationship: logit(germination) = -1.5 + 0.025 * time
    log_odds = -1.5 + 0.025 * stratification_time,
    true_prob = plogis(log_odds),
    # Simulate germination (binomial process)
    n_germinated = rbinom(n(), n_seeds, true_prob),
    n_failed = n_seeds - n_germinated,
    proportion = n_germinated / n_seeds
  )

# Visualize
ggplot(germination_data, aes(x = factor(stratification_time), y = proportion)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(x = "Stratification Time (days)", y = "Proportion Germinated")

# -------------------------------------------
# WRONG: Linear model on proportions
# -------------------------------------------

wrong_model <- lm(proportion ~ stratification_time, data = germination_data)
summary(wrong_model)

# Check residuals — look for heteroscedasticity
par(mfrow = c(2, 2))
plot(wrong_model)

# -------------------------------------------
# CORRECT: Binomial GLM with counts
# -------------------------------------------

correct_model <- glm(cbind(n_germinated, n_failed) ~ stratification_time, 
                     family = binomial, 
                     data = germination_data)
summary(correct_model)

# Check for overdispersion
# Residual deviance / residual df should be approximately 1
resid_dev <- summary(correct_model)$deviance
resid_df <- summary(correct_model)$df.residual
cat("Dispersion ratio:", resid_dev / resid_df, "\n")
# If >> 1, use quasibinomial or beta-binomial

# -------------------------------------------
# Compare predictions
# -------------------------------------------

new_data <- data.frame(stratification_time = 0:90)
new_data$pred_wrong <- predict(wrong_model, new_data)
new_data$pred_correct <- predict(correct_model, new_data, type = "response")

ggplot() +
  geom_point(data = germination_data, 
             aes(x = stratification_time, y = proportion), alpha = 0.5) +
  geom_line(data = new_data, 
            aes(x = stratification_time, y = pred_wrong, color = "Linear Model")) +
  geom_line(data = new_data, 
            aes(x = stratification_time, y = pred_correct, color = "Binomial GLM")) +
  labs(y = "Proportion Germinated", color = "Model") +
  ylim(0, 1)
```

### What students should observe

| Issue | Linear Model | Binomial GLM |
|-------|--------------|--------------|
| Predictions at extremes | Can go below 0 or above 1 | Always between 0 and 1 |
| Residual variance | Heteroscedastic (funnel shape) | Properly modeled |
| Interpretation | Additive effect on proportion | Multiplicative effect on odds |
| Accounts for n_seeds | No | Yes |

**Key insight:** When you have count data out of a known total (successes/trials), use `cbind(successes, failures)` with `family = binomial`. This properly models the binomial nature of the data.

---

# 🔴 Track C: Advanced Stations

## Station 2: Wheat Yield Trials — Answer

### What's wrong with `aov(yield ~ variety)`?

The proposed model ignores the **block structure** of the RCBD (Randomized Complete Block Design). Blocks account for spatial heterogeneity (some blocks may be more fertile).

**Consequences of ignoring blocks:**

- Block variation is lumped into residual error
- Residual variance is inflated
- Standard errors are larger
- F-test has less power
- May miss real treatment effects

### The correct approach

```{r}
#| label: station2-answer
#| eval: false

library(tidyverse)
library(lme4)
library(emmeans)

# -------------------------------------------
# Simulate RCBD data
# -------------------------------------------

set.seed(2024)

varieties <- c("Var_A", "Var_B", "Var_C", "Var_D", "Var_E")
blocks <- c("Block_1", "Block_2", "Block_3", "Block_4")

wheat_data <- expand.grid(
  variety = varieties,
  block = blocks,
  stringsAsFactors = FALSE
) %>%
  mutate(
    # True variety effects (differences from grand mean)
    variety_effect = case_when(
      variety == "Var_A" ~ -200,
      variety == "Var_B" ~ 0,
      variety == "Var_C" ~ 400,  # Best variety
      variety == "Var_D" ~ -100,
      variety == "Var_E" ~ 200
    ),
    # Block effects (some blocks more fertile)
    block_effect = case_when(
      block == "Block_1" ~ -400,
      block == "Block_2" ~ -100,
      block == "Block_3" ~ 200,
      block == "Block_4" ~ 300
    ),
    # Grand mean + effects + residual error
    yield = 5000 + variety_effect + block_effect + rnorm(n(), 0, 120)
  )

# Visualize
ggplot(wheat_data, aes(x = variety, y = yield, color = block)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Wheat Yield by Variety and Block")

# -------------------------------------------
# WRONG: Ignoring blocks
# -------------------------------------------

wrong_model <- aov(yield ~ variety, data = wheat_data)
summary(wrong_model)

# Note: Residual MS includes block variation!

# -------------------------------------------
# CORRECT Option A: Block as fixed effect
# -------------------------------------------

correct_fixed <- aov(yield ~ block + variety, data = wheat_data)
summary(correct_fixed)

# Compare residual variance
cat("Wrong model residual MS:", summary(wrong_model)[[1]]["Residuals", "Mean Sq"], "\n")
cat("Correct model residual MS:", summary(correct_fixed)[[1]]["Residuals", "Mean Sq"], "\n")

# -------------------------------------------
# CORRECT Option B: Block as random effect
# -------------------------------------------

correct_random <- lmer(yield ~ variety + (1 | block), data = wheat_data)
summary(correct_random)
anova(correct_random)

# -------------------------------------------
# Post-hoc comparisons
# -------------------------------------------

emmeans_result <- emmeans(correct_fixed, pairwise ~ variety)
emmeans_result$contrasts

# Compact letter display
library(multcomp)
cld(emmeans_result$emmeans, Letters = letters)

# -------------------------------------------
# Extension: Publication figure
# -------------------------------------------

# Get estimated marginal means
emm <- as.data.frame(emmeans_result$emmeans)

ggplot(emm, aes(x = reorder(variety, emmean), y = emmean)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_jitter(data = wheat_data, aes(x = variety, y = yield), 
              width = 0.1, alpha = 0.4, color = "gray50") +
  labs(x = "Variety", y = "Yield (kg/ha)",
       title = "Wheat Variety Comparison",
       subtitle = "Points = raw data, Error bars = 95% CI of estimated means") +
  theme_minimal()
```

### Fixed vs. Random blocks?

| Use Fixed Block | Use Random Block |
|-----------------|------------------|
| Blocks are exhaustive (these specific blocks) | Blocks are a sample from a population |
| Interest in specific block differences | Want to generalize beyond these blocks |
| Blocks are part of the experimental design | Blocks are nuisance variation |
| Traditional agronomic trials | Multi-site, multi-year studies |

**For this scenario:** Either is defensible. Fixed is more traditional in agronomy; random is more generalizable.

---

## Station 3: Bird Abundance Over Time — Answer

### What are the THREE problems with `lm(bird_count ~ year)`?

1. **Non-normal response:** Counts follow Poisson (or negative binomial), not normal distribution

2. **Non-independence within parks:** Multiple observations from the same park are correlated (need random effect for park)

3. **Temporal pseudoreplication:** Multiple visits within years are not independent

### The correct approach

```{r}
#| label: station3-answer
#| eval: false

library(tidyverse)
library(lme4)
library(glmmTMB)

# -------------------------------------------
# Simulate repeated measures count data
# -------------------------------------------

set.seed(303)

n_parks <- 8
n_years <- 5
n_visits <- 6

bird_data <- expand.grid(
  park_id = 1:n_parks,
  year = 1:n_years,
  visit = 1:n_visits
) %>%
  mutate(
    # Park-level random intercepts
    park_effect = rep(rnorm(n_parks, 0, 0.4), each = n_years * n_visits),
    # Optional: Park-level random slopes (some parks decline faster)
    park_slope = rep(rnorm(n_parks, 0, 0.05), each = n_years * n_visits),
    # True model: log(count) = 3.5 - 0.15*year + park_effect + park_slope*year
    log_mean = 3.5 - 0.15 * year + park_effect + park_slope * year,
    # Simulate Poisson counts
    bird_count = rpois(n(), exp(log_mean))
  )

# Visualize
ggplot(bird_data, aes(x = year, y = bird_count, color = factor(park_id))) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  geom_smooth(method = "glm", method.args = list(family = "poisson"), se = FALSE) +
  facet_wrap(~ park_id, ncol = 4) +
  labs(title = "Bird Counts by Park Over Time")

# -------------------------------------------
# WRONG: Linear model ignoring everything
# -------------------------------------------

wrong_model <- lm(bird_count ~ year, data = bird_data)
summary(wrong_model)

# Problems visible in residuals
par(mfrow = c(2, 2))
plot(wrong_model)

# -------------------------------------------
# BETTER: Poisson GLMM with random intercepts
# -------------------------------------------

better_model <- glmer(bird_count ~ year + (1 | park_id), 
                      family = poisson, 
                      data = bird_data)
summary(better_model)

# Check for overdispersion
overdisp <- sum(residuals(better_model, type = "pearson")^2) / df.residual(better_model)
cat("Overdispersion ratio:", overdisp, "\n")
# If > 1.5, consider negative binomial

# -------------------------------------------
# BEST: Random slopes (parks decline at different rates)
# -------------------------------------------

best_model <- glmer(bird_count ~ year + (1 + year | park_id), 
                    family = poisson, 
                    data = bird_data)
summary(best_model)

# Compare models
anova(better_model, best_model)

# -------------------------------------------
# Interpretation
# -------------------------------------------

# The year coefficient is on the log scale
year_coef <- fixef(best_model)["year"]
cat("Year effect (log scale):", year_coef, "\n")
cat("Multiplicative change per year:", exp(year_coef), "\n")
cat("Percent change per year:", (exp(year_coef) - 1) * 100, "%\n")

# -------------------------------------------
# Extension: Negative binomial for overdispersion
# -------------------------------------------

nb_model <- glmmTMB(bird_count ~ year + (1 + year | park_id), 
                    family = nbinom2, 
                    data = bird_data)
summary(nb_model)
```

### What students should observe

| Model | Issues Addressed |
|-------|------------------|
| `lm(count ~ year)` | None |
| `glm(..., family = poisson)` | Count distribution |
| `glmer(... + (1 \| park))` | Count + park clustering |
| `glmer(... + (1 + year \| park))` | Count + park clustering + varying trends |
| `glmmTMB(..., family = nbinom2)` | All above + overdispersion |

---

## Station 5: Pollinator Network Complexity — Answer

### Why is the Bayesian spatial model overkill?

The proposed method (Bayesian multilevel model with spatial Gaussian process) would require:
- Spatial coordinates (not provided)
- Grouping structure for random effects (not present)
- Substantial computation time
- Expertise in Bayesian methods
- Complex model diagnostics

**For 25 independent meadows with a simple linear relationship, this is massive overkill.**

### The correct approach: Simple linear regression

```{r}
#| label: station5-answer
#| eval: false

library(tidyverse)

# -------------------------------------------
# Simulate simple, well-behaved data
# -------------------------------------------

set.seed(555)

n_meadows <- 25

pollinator_data <- data.frame(
  meadow_id = 1:n_meadows,
  plant_richness = round(runif(n_meadows, 5, 45))
) %>%
  mutate(
    # Simple linear relationship with normal errors
    network_complexity = 1.0 + 0.15 * plant_richness + rnorm(n_meadows, 0, 0.7)
  )

# Visualize
ggplot(pollinator_data, aes(x = plant_richness, y = network_complexity)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Plant Species Richness", 
       y = "Network Complexity Score",
       title = "Simple Linear Relationship")

# -------------------------------------------
# THE CORRECT APPROACH: Simple lm()
# -------------------------------------------

simple_model <- lm(network_complexity ~ plant_richness, data = pollinator_data)
summary(simple_model)

# Check assumptions
par(mfrow = c(2, 2))
plot(simple_model)

# All assumptions met! We're done.

# -------------------------------------------
# Discussion: When WOULD complexity be needed?
# -------------------------------------------

# The complex approach would be appropriate if:

# 1. Meadows were grouped (e.g., multiple meadows per region)
#    → Need random intercepts for region

# 2. Repeated measures over time
#    → Need random intercepts for meadow, possibly temporal autocorrelation

# 3. Spatial autocorrelation (nearby meadows more similar)
#    → Need spatial covariance structure (Gaussian process, CAR, etc.)

# 4. Multiple response variables
#    → Need multivariate model

# 5. Prior information from previous studies
#    → Bayesian framework with informative priors

# -------------------------------------------
# Costs of over-complication
# -------------------------------------------

# 1. Computation time (hours vs. seconds)
# 2. More parameters to estimate with limited data (n=25)
# 3. Harder to interpret and explain
# 4. More opportunities for errors
# 5. Reviewers may not understand or trust the analysis
# 6. May not converge or give unreliable estimates
# 7. Requires specialized software and expertise
```

### Key discussion points

**Occam's Razor in Statistics:** Start with the simplest model that addresses your research question. Add complexity only when:

1. The simple model clearly fails (poor fit, violated assumptions)
2. The research question requires it (e.g., estimating variance components)
3. The data structure demands it (e.g., hierarchical, spatial, temporal)

**Red flags for unnecessary complexity:**

- "I used a Bayesian spatial model" for simple cross-sectional data
- Random effects with < 5 groups
- More parameters than you can reliably estimate
- Methods chosen to seem sophisticated rather than answer the question

---

# General Discussion Points for Wrap-Up

## Why simulate data?

1. **Forces you to understand the data-generating process** — You can't simulate what you don't understand
2. **Creates known "truth" to compare methods against** — You know the true effect size
3. **Reveals how methods fail** — See the consequences of wrong choices
4. **Builds intuition** — Understand what different parameters do
5. **Useful for power analysis** — Before collecting real data

## Common themes across stations

| Station | Main Issue | Lesson |
|---------|------------|--------|
| 1 (Salamander) | Pseudoreplication | Know your unit of replication |
| 2 (Wheat) | Ignoring design structure | Use the structure you designed |
| 3 (Birds) | Multiple violations | Match method to data structure |
| 4 (Soil) | Overcautious | Don't be afraid of parametric tests |
| 5 (Pollinator) | Overkill | Simpler is often better |
| 6 (Germination) | Wrong distribution | Match distribution to response type |

## Signs you've chosen the wrong method

- Residual plots look terrible
- Predictions outside possible range
- Standard errors that seem too small (pseudoreplication)
- Model won't converge
- Results don't make biological sense
- You can't explain what the model is doing

</div>

<script>
function checkPassword() {
  const password = document.getElementById('password-input').value;
  
  // =============================================
  // CHANGE THIS PASSWORD TO YOUR DESIRED PASSWORD
  // =============================================
  const correctPassword = "SNR690answers";
  
  if (password === correctPassword) {
    document.getElementById('password-form').style.display = 'none';
    document.getElementById('protected-content').style.display = 'block';
    sessionStorage.setItem('authenticated', 'true');
  } else {
    document.getElementById('error-msg').textContent = 'Incorrect password. Try again.';
    document.getElementById('password-input').value = '';
  }
}

// Check if already authenticated this session
if (sessionStorage.getItem('authenticated') === 'true') {
  document.getElementById('password-form').style.display = 'none';
  document.getElementById('protected-content').style.display = 'block';
}

// Allow Enter key to submit
document.getElementById('password-input').addEventListener('keypress', function(e) {
  if (e.key === 'Enter') {
    checkPassword();
  }
});
</script>